The ChEMBL database as Linked Open Data

### 2012-05-28 article.tex
Small tweakings

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL datadump provided by EMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences XXX Working Group~\cite{Marshall2012}.

@@

\begin{table*}
\caption{Prefixes and their matching namespaces used in this paper.} \label{namespaces}
\begin{tabular}{ll}
\hline
\multicolumn{2}{l}{\textbf{Common Vocabularies}} \\
bibo    & Bibliography Ontology~\cite{Giasson2011} \\
        & http://purl.org/ontology/bibo/ \\
chebi   & Chemical Entities of Biological Interest~\cite{DeMatos2010} \\
        & http://purl.org/obo/owl/CHEBI\# \\
cheminf & Chemical Information Ontology~\cite{Hastings2011} \\
        & http://semanticscience.org/resource/ \\
cito    & Citation Typing Ontology~\cite{Shotton2010} \\
        & http://purl.org/spar/cito/ \\
obo / pro & OBO \& PRotein Ontology~\cite{Sidhu2006} \\
          & http://purl.obolibrary.org/obo/ \\

@@

\multicolumn{2}{l}{\textbf{ChEMBL-RDF Prefixes }}\\
act    & http://data.kasabi.com/dataset/chembl-rdf/activity/ \\
assay  & http://data.kasabi.com/dataset/chembl-rdf/assay/ \\
mol    & http://data.kasabi.com/dataset/chembl-rdf/molecule/ \\
res    & http://data.kasabi.com/dataset/chembl-rdf/resource/ \\
\hline
\end{tabular}
\end{table*}

@@

\begin{figure}[t]
\includegraphics[width=0.45\textwidth]{figs/relations}
\caption{The various resource types found in the ChEMBL triples. Some entities are subclasses
of common classes, while others are instances.}\label{f1}
\end{figure}

@@

\begin{footnotesize}
\begin{verbatim}
act:a31863
 a       chembl:Activity ;
 cito:citesAsDataSource
  res:r6424 ;
 chembl:forMolecule mol:m180094 ;
 chembl:onAssay assay:a54505 ;
 chembl:relation ">" ;
 chembl:standardUnits
  "nM" ;
 chembl:standardValue
  "100000"^^xsd:float ;
 chembl:type "IC50" .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
assay:a17
 a chembl:Assay ;
 cito:citesAsDataSource
  res:r11347 ;
 chembl:hasAssayType chembl:ADMET ;
 chembl:hasConfScore "7"^^xsd:int ;
 chembl:hasDescription
  "Inhibition of ... hydroxylase" ;
 chembl:hasTarget target:t100122 .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
target:t1
 a chembl:Target ;
 rdfs:subClassOf pro:PR_000000001 ;
 rdfs:label "Glucoamylase" , 
  "Maltase-glucoamylase, intestinal" ;
 dc:identifier "uniprot:O43451" ,
   "3.2.1.3" ;
 dc:title "Maltase-glucoamylase" ;
 chembl:classL1 "Enzyme" ;
 chembl:hasDescription
  "Maltase-glucoamylase, intestinal" ;
 chembl:hasKeyword "Glycosidase" , 
  "Membrane" , "Sulfation" ;
 chembl:organism "Homo sapiens" .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
mol:m4 obo:has_role chebi:CHEBI_23888 .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
mol:m41
 rdfs:label "ChEMBL406142" , 
   "Bis(3-[1 .... yl]-propionamide)" ;
 rdfs:subClassOf cheminf:CHEMINF_000000 ;
 cheminf:CHEMINF_000200
  m41:inchikey , m41:smiles , m41:inchi ;

@@

chemblid:ChEMBL406142
 owl:equivalentClass mol:m41 .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
mol:m1
 cheminf:CHEMINF_000200 m1:alogp .

m1:alogp
 a cheminf:CHEMINF_000305 ;
 cheminf:SIO_000300
  "3.344"^^xsd:double .
\end{verbatim}
\end{footnotesize}

@@

\begin{footnotesize}
\begin{verbatim}
journal:j6c706049c2e08871b7c46a6528065736
 a bibo:Journal ;
 dc:title "J. Med. Chem." .

res:r1
 a bibo:Article ;
 dc:date "2004" ;
 dc:isPartOf
  journal:j6c706049c2e08871b7c46a6528065736 ;
 bibo:issue "1" ;
 bibo:pageEnd "9" ;
 bibo:pageStart "1" ;
 bibo:pmid "14695813" ;
 bibo:volume "47" .
\end{verbatim}
\end{footnotesize}

@@

\begin{tiny}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p bibo:Article .
}
\end{verbatim}
\end{tiny}

@@

\begin{figure}[t]
\includegraphics[width=0.45\textwidth]{figs/lodgraph}
\caption{The links out of the ChEMBL-RDF data into the Linked Open Data cloud.
Edges are labeled by the predicates making the links.}\label{2}
\end{figure}

>>>>

The current scientific data deluge, in which datasets are growing faster than scientists can analyse and 
curate them, provides several new challenges for information systems. Scientists wish to 
discover new, unique, and significant patterns in datasets, which explaining biological 
phenomena not yet understood. This process requires integration of datasets that are both 
sparse, and growing independent of one another. The discovery of new patterns showing the 
causes and effects of various biological phenomena is beyond the scope of single datasets.
For example, systems biologists integrate micro-array differential expression datasets to 
biological pathways, using various other datasets to provide evidence for the links~\cite{FIXME}. 
Another prominent example is drug discovery where a new unique chemical entity is discovered 
based on descriptions of its biological properties. This process requires the effective 
linkage of many scientific datasets~\cite{Samwald2011,OpenPHACTS}.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
The SQL datadump provided by ChEMBL was inserted into a local MySQL database. A set of SQL queries 
were then executed to generate RDF files using a set of PHP scripts. These scripts are Open Source 
and are available from the source code hosting service, GitHub~\cite{ChEMBLRDFGitHub}. This process 
has been outlined in a best practices note by the W3C Health Care and Life Sciences Interest Group~\cite{Marshall2012}.

@@

\begin{table*}
\caption{Prefixes and their matching namespaces used in this paper.} \label{namespaces}
\begin{center}
\begin{tabular}{ll}
\hline
\multicolumn{2}{l}{\textbf{Common Vocabularies}} \\
bibo    & Bibliography Ontology~\cite{Giasson2011} \\
        & http://purl.org/ontology/bibo/ \\
chebi   & Chemical Entities of Biological Interest~\cite{DeMatos2010} \\
        & http://purl.org/obo/owl/CHEBI\# \\
cheminf & Chemical Information Ontology~\cite{Hastings2011} \\
        & http://semanticscience.org/resource/ \\
cito    & Citation Typing Ontology~\cite{Shotton2010} \\
        & http://purl.org/spar/cito/ \\
obo / pro & OBO \& PRotein Ontology~\cite{Sidhu2006} \\
          & http://purl.obolibrary.org/obo/ \\

@@

\multicolumn{2}{l}{\textbf{ChEMBL-RDF Prefixes }}\\
act    & http://data.kasabi.com/dataset/chembl-rdf/activity/ \\
assay  & http://data.kasabi.com/dataset/chembl-rdf/assay/ \\
mol    & http://data.kasabi.com/dataset/chembl-rdf/molecule/ \\
res    & http://data.kasabi.com/dataset/chembl-rdf/resource/ \\
\hline
\end{tabular}
\end{center}
\end{table*}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=0.7\textwidth]{figs/relations}
\caption{The various resource types found in the ChEMBL triples. Some entities are subclasses
of common classes, while others are instances.}\label{f1}
\end{center}
\end{figure}

@@

\begin{small}
\begin{verbatim}
act:a31863
 a       chembl:Activity ;
 cito:citesAsDataSource
  res:r6424 ;
 chembl:forMolecule mol:m180094 ;
 chembl:onAssay assay:a54505 ;
 chembl:relation ">" ;
 chembl:standardUnits
  "nM" ;
 chembl:standardValue
  "100000"^^xsd:float ;
 chembl:type "IC50" .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
assay:a17
 a chembl:Assay ;
 cito:citesAsDataSource
  res:r11347 ;
 chembl:hasAssayType chembl:ADMET ;
 chembl:hasConfScore "7"^^xsd:int ;
 chembl:hasDescription
  "Inhibition of ... hydroxylase" ;
 chembl:hasTarget target:t100122 .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
target:t1
 a chembl:Target ;
 rdfs:subClassOf pro:PR_000000001 ;
 rdfs:label "Glucoamylase" , 
  "Maltase-glucoamylase, intestinal" ;
 dc:identifier "uniprot:O43451" ,
   "3.2.1.3" ;
 dc:title "Maltase-glucoamylase" ;
 chembl:classL1 "Enzyme" ;
 chembl:hasDescription
  "Maltase-glucoamylase, intestinal" ;
 chembl:hasKeyword "Glycosidase" , 
  "Membrane" , "Sulfation" ;
 chembl:organism "Homo sapiens" .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
mol:m4 obo:has_role chebi:CHEBI_23888 .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
mol:m41
 rdfs:label "ChEMBL406142" , 
   "Bis(3-[1 .... yl]-propionamide)" ;
 rdfs:subClassOf cheminf:CHEMINF_000000 ;
 cheminf:CHEMINF_000200
  m41:inchikey , m41:smiles , m41:inchi ;

@@

chemblid:ChEMBL406142
 owl:equivalentClass mol:m41 .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
mol:m1
 cheminf:CHEMINF_000200 m1:alogp .

m1:alogp
 a cheminf:CHEMINF_000305 ;
 cheminf:SIO_000300
  "3.344"^^xsd:double .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
journal:j6c706049c2e08871b7c46a6528065736
 a bibo:Journal ;
 dc:title "J. Med. Chem." .

res:r1
 a bibo:Article ;
 dc:date "2004" ;
 dc:isPartOf
  journal:j6c706049c2e08871b7c46a6528065736 ;
 bibo:issue "1" ;
 bibo:pageEnd "9" ;
 bibo:pageStart "1" ;
 bibo:pmid "14695813" ;
 bibo:volume "47" .
\end{verbatim}
\end{small}

@@

\begin{small}
\begin{verbatim}
select count(*) WHERE {
  ?s ?p bibo:Article .
}
\end{verbatim}
\end{small}

@@

\begin{figure}[t]
{\center
\includegraphics[width=0.45\textwidth]{figs/lodgraph}
\caption{The links out of the ChEMBL-RDF data into the Linked Open Data cloud.
Edges are labeled by the predicates making the links.}\label{2}
}
\end{figure}
